---
title: "SAMprior for Continuous Endpoints"
author: "Peng Yang and Ying Yuan"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
  html_vignette:
    toc: true
  html_document:
    toc: true
    number_sections: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  pdf_document:
    toc: true
  word_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with SAMprior (continuous)}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---
  
```{r, include=FALSE}
library(SAMprior)
library(knitr)
knitr::opts_chunk$set(
    fig.width = 1.62*4,
    fig.height = 4
    )
## setup up fast sampling when run on CRAN
is_CRAN <- !identical(Sys.getenv("NOT_CRAN"), "true")
## NOTE: for running this vignette locally, please uncomment the
## following line:
## is_CRAN <- FALSE
.user_mc_options <- list()
if (is_CRAN) {
    .user_mc_options <- options(RBesT.MC.warmup=250, RBesT.MC.iter=500, RBesT.MC.chains=2, RBesT.MC.thin=1, RBesT.MC.control=list(adapt_delta=0.9))
}
```

# Introduction

The self-adapting mixture prior (SAMprior) package is designed to enhance 
the effectiveness and practicality of clinical trials by leveraging historical 
information or real-world data [1]. The package incorporate historical data 
into a new trial using an informative prior constructed based on historical 
data while mixing a non-informative prior to enhance the robustness of 
information borrowing. It utilizes a data-driven way to determine a self-adapting
mixture weight that dynamically favors the informative (non-informative) prior
component when there is little (substantial) evidence of prior-data conflict.
Operating characteristics are evaluated and compared to the robust 
Meta-Analytic-Predictive (rMAP) prior [2], which assigns a fixed weight of 0.5. 

# SAM Prior Derivation

SAM prior is constructed by mixing an informative prior $\pi_1(\theta)$, 
constructed based on historical data, with a non-informative prior 
$\pi_0(\theta)$ using the mixture weight $w$ determined by **`SAM_weight`**
function to achieve the degree of prior-data conflict [1]. The following 
sections describe how to construct SAM prior in details.

## Informative Prior Construction based on Historical Data

We assume three historical data as follows:
```{r,results="D_h",echo=FALSE}
set.seed(123)
std <- function(x) sd(x)/sqrt(length(x))
df_1 <- rnorm(40, 0, 3); 
df_2 <- rnorm(50, 0, 3); 
df_3 <- rnorm(60, 0, 3); 
dat <- data.frame(study = c(1,2,3),
                  n = c(40, 50, 60),
                  mean = round(c(mean(df_1), mean(df_2), mean(df_3)), 3),
                  se = round(c(std(df_1), std(df_2), std(df_3)), 3))
kable(dat)
```



To construct informative priors based on the aforementioned three historical 
data, we apply **`gMAP`** function from RBesT to perform meta-analysis. This 
informative prior results in a representative form from a large MCMC samples, 
and it can be converted to a parametric representation with the 
**`automixfit`** function using expectation-maximization (EM) algorithm [3].
This informative prior is also called MAP prior.

```{r, message=FALSE}
sigma = 3
# load R packages
library(ggplot2)
theme_set(theme_bw()) # sets up plotting theme
set.seed(22)
map_mcmc <- gMAP(cbind(mean, se) ~ 1 | study, 
                 weights=n,data=dat,
                 family=gaussian,
                 beta.prior=cbind(0, sigma),
                 tau.dist="HalfNormal",tau.prior=cbind(0,sigma/2))

map_automix <- automixfit(map_mcmc)
map_automix
plot(map_automix)$mix
```

The resulting MAP prior is approximated by a mixture of conjugate priors.

## SAM Weight Determination

Let $\theta$ and $\theta_h$ denote the treatment effects associated with the
current arm data $D$ and historical $D_h$, respectively. Let $\delta$ denote
the clinically significant difference such that is $|\theta_h - \theta| \ge \delta$,
then $\theta_h$ is regarded as clinically distinct from $\theta$, and it is 
therefore inappropriate to borrow any information from $D_h$. Consider two
hypotheses:

$$
H_0: \theta = \theta_h, ~~ H_1: \theta = \theta_h + \delta ~ \text{or} ~ \theta = \theta_h - \delta.
$$
$H_0$ represents that $D_h$ and $D$ are consistent (i.e., no prior-data 
conflict) and thus information borrowing is desirable, whereas $H_1$ represents
that the treatment effect of $D$ differs from $D_h$ to such a degree that no
information should be borrowed.

The SAM prior uses the likelihood ratio test (LRT) statistics $R$ to quantify
the degree of prior-data conflict and determine the extent of information 
borrowing.
$$
R = \frac{P(D | H_0, \theta_h)}{P(D | H_1, \theta_h)} = \frac{P(D | \theta = \theta_h)}{\max \{ P(D | \theta = \theta_h + \delta), P(D | \theta = \theta_h - \delta) \}} ,
$$
where $P(D | \cdot)$ denotes the likelihood function. An alternative Bayesian
choice is the posterior probability ratio (PPR):
$$
R = \frac{P(D | H_0, \theta_h)}{P(D | H_1, \theta_h)} = \frac{P(H_0)}{P(H_1)} \times BF ,
$$
where $P(H_0)$ and $P(H_1)$ is the prior probabilities of $H_0$ and $H_1$ 
being true. $BF$ is the Bayes Factor that in this case is the same as LRT.

The SAM prior, denoted as $\pi_{sam}(\theta)$, is then defined as a mixture
of an informative prior $\pi_1(\theta)$, constructed based on $D_h$, with a 
non-informative prior $\pi_0(\theta)$:
$$\pi_{sam}(\theta) = w \pi_1(\theta) + (1 - w) \pi_0(\theta)$$
where the mixture weight $w$ is calculated as:
$$w = \frac{R}{1 + R}.$$ 
As the level of prior-data conflict increases, the likelihood ratio $R$ 
decreases, resulting in a decrease in the weight $w$ assigned to the 
informative prior and a decrease in information borrowing. As a result,
$\pi_{sam}(\theta)$ is data-driven and has the ability to self-adapt the 
information borrowing based on the degree of prior-data conflict.

To calculate the SAM weight $w$, we first assume the sample size enrolled in 
the control arm is $n = 30$, with $\theta = 0.4$ and $\sigma = 3$. Additionally,
we assume the effective size is $d = \frac{\theta - \theta_h}{\sigma} = 0.5$,
then we can apply 
function **`SAM_weight`** in SAMprior as follows: 
```{r, message=FALSE}
set.seed(234)
sigma        <- 3 ## Standard deviation in the current trial
data.crt     <- rnorm(30, mean = 0.4, sd = sigma)
wSAM <- SAM_weight(if.prior = map_automix, 
                   delta = 0.5 * sigma,
                   data = data.crt)
cat('SAM weight: ', wSAM)
```

The default method to calculate $w$ is using LRT, which is fully data-driven.
However, if investigators want to incorporate prior information on prior-data
conflict to determine the mixture weight $w$, this can be achieved by using 
PPR method as follows:
```{r, message=FALSE}
wSAM <- SAM_weight(if.prior = map_automix, 
                   delta = 0.5 * sigma,
                   method.w = 'PPR',
                   prior.odds = 3/7,
                   data = data.crt)
cat('SAM weight: ', wSAM)
```
The **`prior.odds`** indicates the prior probability of $H_0$ over the prior 
probability of $H_1$. In this case (e.g., **`prior.odds = 3/7`**), the prior 
information favors the presence prior-data conflict and it results in a 
decreased mixture weight.


When historical information is congruent with the current control arm, SAM 
weight reaches to the highest peak. As the level of prior-data conflict 
increases, SAM weight decreases. This demonstrates that SAM prior is data-driven
and self-adapting, favoring the informative (non-informative) prior component 
when there is little (substantial) evidence of prior-data conflict.
```{r, echo=FALSE, message=FALSE, warning=FALSE}
weight_grid <- seq(-3, 3, by = 0.3)
weight_res  <- lapply(weight_grid, function(x){
  res <- c()
  for(s in 1:300){
    data.control <- rnorm(n = 35, mean = x, sd = sigma)
    res <- c(res, SAM_weight(if.prior = map_automix,
                             delta = 0.5 * sigma,
                             data = data.control))
    
  }
  mean(res)
})
df_weight <- data.frame(grid   = weight_grid,
                        weight = unlist(weight_res))
qplot(grid, weight, data = df_weight, geom = "line", main= "SAM Weight") +
  xlab('Sample mean from control trial')+ ylab('Weight') +
  geom_vline(xintercept = summary(map_automix)['mean'], linetype = 2, col = 'blue') 
```



## SAM Prior Construction

To construct the SAM prior, we mix the derived informative prior $\pi_1(\theta)$ 
with a vague prior $\pi_0(\theta)$ using pre-determined mixture weight by 
**`SAM_prior`** function in SAMprior as follows: 
```{r, message=FALSE}
unit_prior <- mixnorm(nf.prior = c(1, summary(map_automix)['mean'], sigma))
SAM.prior <- SAM_prior(if.prior = map_automix, 
                       nf.prior = unit_prior,
                       weight = wSAM, sigma = sigma)
SAM.prior
```
where the non-informative prior $\pi_0(\theta)$ follows an unit-information prior.

## Operating Characteristics

In this section, we aim to investigate the operating characteristics of 
the SAM prior, constructed based on the historical data, via simulation.
The incorporation of historical information is expected to be beneficial 
in reducing the required sample size for the current arms. To achieve this, 
we assume a 1:2 ratio between the control and treatment arms.

We compare the operating characteristics of the SAM prior and rMAP prior 
with pre-specified fixed weight under various scenarios. Specifically, 
we will evaluate the relative bias and relative mean squared error (MSE) 
of these methods. The relative bias and relative MSE are defined 
as the differences between the bias/MSE of a given method and the bias/MSE 
obtained when using a non-informative prior.

Additionally, we investigate the type I error and power of the methods 
under different degrees of prior-data conflicts. The decision regarding 
whether a treatment is superior or inferior to a standard control will be 
based on the condition:
$$\Pr(\theta_t - \theta > 0) > 0.95.$$

In SAMprior, the operating characteristics can be considered in following steps:

1. Specify priors: This step involves constructing informative prior based 
   on historical data and non-informative prior.

2. Specify the decision criterion: The **`decision2S`** function 
   is used to initialize the decision criterion. This criterion determines 
   whether a treatment is considered superior or inferior to a standard 
   control.
   
3. Specify design parameters for the **`get_OC`** function: This step 
   involves defining the design parameters for evaluating the operating 
   characteristics. These parameters include the clinically significant 
   difference (CSD) used in SAM prior calculation, the method used to 
   determine the mixture weight for the SAM prior, the sample sizes for 
   the control and treatment arms, the number of trials used for simulation, 
   the choice of weight for the robust MAP prior used as a benchmark, and 
   the vector of mean for both the control and treatment arms.

### Type I Error

To compute the type I error, we consider four scenarios, with the first and 
last two scenarios representing minimal and substantial prior-data conflicts, 
respectively. In general, the results show that both methods effectively 
control the type I error.
```{r, message=FALSE}
set.seed(123)
# weak_prior <- mixnorm(c(1, summary(map_automix)[1], 1e4))
TypeI <- get_OC(if.prior = map_automix,    ## MAP prior from historical data
                nf.prior = unit_prior,     ## Weak-informative prior for treatment arm
                delta    = 0.5*sigma,      ## CSD for SAM prior
                n        = 35, n.t = 70,   ## Sample size for control and treatment arms
                ## Decisions
                decision = decision2S(0.95, 0, lower.tail=FALSE), 
                ntrial   = 1000,           ## Number of trials simulated
                if.MAP   = TRUE,           ## Output robust MAP prior for comparison
                weight   = 0.5,            ## Weight for robust MAP prior
                ## Mean for control and treatment arms
                theta    = c(0, 0,    -2, 4),
                theta.t  = c(0, -0.1, -2, 4),
                sigma    = sigma
                  )
kable(TypeI)
```
### Power

For power evaluation, we also consider four scenarios, with the first and 
last two scenarios representing minimal and substantial prior-data conflicts, 
respectively. In general, it is observed that the SAM prior achieves comparable 
or better power compared to rMAP.
```{r, message=FALSE}
set.seed(123)
Power <- get_OC(if.prior = map_automix,    ## MAP prior based on historical data
                nf.prior = unit_prior,     ## Non-informative prior for treatment arm
                delta    = 0.5*sigma,      ## CSD for SAM prior
                n        = 35, n.t = 70,   ## Sample size for control and treatment arms
                ## Decisions
                decision = decision2S(0.95, 0, lower.tail=FALSE), 
                ntrial   = 1000,           ## Number of trials simulated
                if.MAP   = TRUE,           ## Output robust MAP prior for comparison
                weight   = 0.5,            ## Weight for robust MAP prior
                ## Mean for control and treatment arms
                theta    = c(0, 0.1, 0.5,  -3),
                theta.t  = c(1, 1.1, 2.0,  -1.5),
                sigma    = sigma
                  )
kable(Power)
```


# Decision Making

Finally, we present an example of how to make a final decision on whether the 
treatment is superior or inferior to a standard control once the trial has been 
completed and data has been collected. This step can be accomplished using the 
**`postmix`** function from RBesT, as shown below:
```{r, message=FALSE}
## Simulate data for treatment arm
data.trt <- rnorm(60, mean = 3, sd = sigma)

## first obtain posterior distributions...
post_SAM <- postmix(priormix = SAM.prior,   ## SAM Prior
                    data = data.crt)
post_trt <- postmix(priormix = unit_prior,  ## Non-informative prior
                    data = data.trt)

## Define the decision function
decision = decision2S(0.95, 0, lower.tail=FALSE)

## Decision-making
decision(post_trt, post_SAM)
```

### References

[1] Yang P. et al., _Arxiv_, https://arxiv.org/abs/2305.12279. \
[2] Schmidli H. et al., _Biometrics_ 2014;70(4):1023-1032. \
[3] Neuenschwander B. et al., _Clin Trials_. 2010; 7(1):5-18. 




### R Session Info

```{r}
sessionInfo()
```

```{r,include=FALSE}
options(.user_mc_options)
```
